---
title: "MY470 - Class Week 9 - More R Programming"
author: "Pablo Barbera"
---

### Working with data frames

#### Factorizing a categorical variable

Using the `mtcars` dataset, you need to transform the `am` variable into a factor with two levels, automatic and manual.  Create a new column of `mtcars` called `am_factor` that is such a factor variable.

```{r}
data(mtcars) # this line of code will load the data into your workspace

mtcars$am_factor <- factor(ifelse(mtcars$am==1, "Automatic", "Manual"))
```

#### Computing aggregated statistics

Create a table with the median value of `mpg` for each value of `am_factor` using the `aggregate` function. Read the help page for `aggregate` to find information on how to achieve this.

```{r}
aggregate(mpg ~ am_factor, data=mtcars, FUN=median)
```

Now replicate the code above, but this time with the average instead of the median.

```{r}
aggregate(mpg ~ am_factor, data=mtcars, FUN=mean)
```

### Working with functions

Write an R function that checks if a string is a palindrome. As you may recall from earlier in the term, a palindrome is a word or phrase that reads the same backward as forward. Note that you may run into some of the limitations of R when working with strings. 

```{r}
is.palindrome <- function(string){
  # split into characters
  chars <- strsplit(string, split="")
  # convert to vector
  chars <- unlist(chars)
  # reverse characters
  rev_chars <- chars[length(chars):1]
  # paste into a single word
  rev_word <- paste(rev_chars, collapse="")
  return(
    ifelse(string==rev_word, "Is palindrome", "Is not palindrome")
  )
}

is.palindrome("racecar")
is.palindrome("level")
is.palindrome("programming")
```


### Loops

Loops in R have a similar structure as in python. For example:

```{r}
for (i in 1:10){
	print(i)
}
```

Write a loop to compute a [Fibonacci sequence](https://www.mathsisfun.com/numbers/fibonacci-sequence.html) in R for an input of length `n`.  We realize that the Internet is full of solutions for this, and you are free to use one, but you should discuss why you chose this solution and what advantages it has over alternatives.

Compute and display the Fibonacci sequence of length 20 using your function.

```{r}
n <- 20

fibonacci <- function(n){
  fib <- c(0, 1, NA)
  for (i in 3:n){
    fib[i] <- fib[i-1] + fib[i-2]
  }
  return(fib)
}

fibonacci(n=20)

```




