---
title: "MY470 - Exercise Week 8 - Programming"
author: "Pablo Barberá"
---

The goal of this exercise is to practice writing code in R. You will have to use the content from this week's lecture as well as the required readings and the R documentation in order to answer the questions below.

We will mark the assignment based on the compiled html file, so please check that the final version of your submission includes that file. Submissions without a compiled html file will receive a mark of 0.

### 1. Working with text

Here, we revisit some of the same issues as in [Week 2](https://github.com/lse-my470/assignment-2/blob/master/MY470_wk2_assign.ipynb).

We will use the same text:
```{r}
txt <- "Someone must have been telling lies about Josef K., he knew he had
done nothing wrong but, one morning, he was arrested.  Every day at
eight in the morning he was brought his breakfast by  Mrs. Grubach's
cook - Mrs. Grubach was his landlady - but today she didn't come.  That
had never happened before.  K. waited a little while, looked from his
pillow at the old woman who lived opposite and who was watching him with
an inquisitiveness quite unusual for her, and finally, both hungry and
disconcert"

# this splits the text by whitespace
toks <- unlist(strsplit(txt, "\\s"))
```

Answer the questions below by creating your own chunks of code to find the right answer:

A) What is the type of the object `toks`?  (Hint: Examine it using the same tools as from the lecture notes.)

*The object `toks` is of class "character"*

```{r}
typeof(toks)
class(toks)
```


B) How many elements are in `toks`? And how many _unique_ elements?

*There are 93 elements, and 74 of them are unique.*

```{r}
length(toks)
length(unique(toks))
```

C) What are the most common three words?  Hint: See `?table`.

*The three most common words are was, he, and his.*

```{r}
# delete punctuation
toks <- gsub("[[:punct:]]", "", toks)
# convert to lower case
toks <- tolower(toks)
# deleting empty words
toks <- toks[toks!=""]

tail(sort(table(toks)))
```

D) Try this using the **quanteda** package.  Functions to perform the above steps are `tokens()`, `ntoken()`, `ntype()`, and `topfeatures`.

*Results are different from our code above because quanteda treats punctuation differently.*

```{r}
library(quanteda)
toks <- tokens(txt)
ntoken(toks)
ntype(toks)
topfeatures(dfm(toks))
```

*Here's how to obtain the same results with our code above and with quanteda.*

```{r}
# with base R:
# 1) convert to lowercase
new_txt <- tolower(txt)
# 2) remove line breaks
new_txt <- gsub("\\n", " ", new_txt)
# 3) removing punctuation
new_txt <- gsub("[[:punct:]]", "", new_txt)
# 4) removing double splaces
new_txt <- gsub("  ", " ", new_txt)

# and now we run the code above...
toks <- unlist(strsplit(new_txt, "\\s"))
length(toks)
length(unique(toks))

# and we get the same as with quanteda
toks <- tokens(txt, remove_punct=TRUE)
ntoken(toks)
ntype(toks)
topfeatures(dfm(toks))
```


### 2. Working with a dataset.

This exercise relates to the `College` data set, which comes from [An Introduction to Statistical Learning](http://www-bcf.usc.edu/~gareth/ISL/data.html) by James et al 2013. It contains a number of variables for 777 different universities and colleges in the US.

For convenience, we have placed the `.csv` file in this repository.

Use the `read.csv()` function to read the data into `R`. Call the loaded data `college`. Make sure that you have the directory set to the correct location for the data.  You can load this in R directly from the website, using:

```{r}
college <- read.csv("http://www-bcf.usc.edu/~gareth/ISL/College.csv")
```

Or you can load it from a saved file, using (you might need to modify the path):

```{r, eval = FALSE}
college <- read.csv("College.csv")
```

A) Look at the data using the `View()` function.  Create a code block and output the results of printing the first 6 rows of the dataset, using the `head()` function.

```{r}
head(college)
```

B) Use the `str()` function to look at the structure of the data. Which of the variables are numeric? Which are integer? Which are factors?

```{r}
str(college)
```

C) Use the `summary()` function to produce a numerical summary of the variables in the data set.

```{r}
summary(college)
```

D) What is the mean and standard deviation of the `Enroll` and `Top10Perc` variables?  You should use the functions `mean()` and `sd()` for the answer.

```{r}
mean(college$Enroll)
sd(college$Enroll)
mean(college$Top10perc)
sd(college$Top10perc)
```

E) Now remove the 10th through 85th observations. This will involve slicing the data, similar to in Python, but for the row element. What is the mean and standard deviation of the `Enroll` and `Top10Perc` variables in the subset of the data that remains?

```{r}
mean(college$Enroll[-(10:85)])
sd(college$Enroll[-(10:85)])
mean(college$Top10perc[-(10:85)])
sd(college$Top10perc[-(10:85)])
# or:
college_subset <- college[-(10:85),]
mean(college_subset$Enroll)
sd(college_subset$Enroll)
mean(college_subset$Top10perc)
sd(college_subset$Top10perc)
```
   
F) What is the range of the `Books` variable?  Hint: Use `range()`.

```{r}
range(college$Books)
```

G) Produce a scatterplot matrix of the first five numeric variables of the dataset.
  
```{r}
library(ggplot2)
library(GGally)
ggpairs(college[,c("Apps", "Accept", "Enroll", "Top10perc", "Top25perc")])
```
